Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-09-02 jso.qmd
done revising introduction

Projection Pursuit (PP) (@kr69, @FT74, @huber85) is a dimension reduction technique aimed at identifying informative linear projections of data. This is useful for exploring high-dimensional data, and creating plots of the data for publication that reveal the main features.  The method involves optimising an objective function known as the PP index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the criteria for what constitutes interesting or informative projections. Let $X \in \mathbb{R}^{n\times p}$ be the data matrix, $A \in\mathbb{R}^{p \times d}$ be an orthonormal matrix, where $A$ belongs to the Stiefel manifold $\mathcal{A} = V_d(\mathbb{R}^p)$. (It is interesting to note that when using PP visually, one cares less about $A$ than the plane described by $A$, because the orientation in the plane is irrelevant. The the space of planes belongs to a Grassmann manifold.) The projection $Y = XA$ is a linear transformation that maps data from a $p$-dimensional space into a $d$-dimensional space. The index function $f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}$ is a scalar function that measures an interesting aspect of the projected data, such as deviation from normality, presence of clusters, non-linear structure, or other features of interest. For a fixed data, PP finds the orthonormal basis $A$ that maximises the index value of the projection, $Y = XA$:

@@

The index functions vary in structure, partly depending on the data that is being projected. @fig-example-functions shows two examples. Huber plots [@huberplot] of 2D data sets are in (a) and (c). The PP index values for all 1D projections of the 2D data are shown in polar coordinates, revealing the nature of the function. The dashed circle is a baseline set at the average value, and the straight line marks the optimal projection. Plots (b) and (d) show the respective best projections of the data as histograms. Indexes like the holes, central mass and skewness [@cook1993projection] are generally smooth for most data sets. Many indexes noisy and non-convex, requiring an effective and efficient optimisation procedure to explore the data landscape and achieve a globally optimal viewpoint of the data. The skewness index computed for trimodal data, in (a), is smooth with a large squint angle but has three modes, and thus is not convex. The binned normality index computed on the famous RANDU data, in (c), is noisier and has a very small squint angle. The discreteness cannot be seen unless the optimiser is very close to the optimal projection.

@@

```{r}
#| label: fig-example-functions
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 80%
#| fig-cap: "Examples of PP indexes with large (top row) and small (bottom row) squint angles, shown with a Huber plot, and histogram of the projected data corresponding to the optimal projection. Huber plot shows PP index values for all 1D projections in polar coordinates."
data(randu)
randu_std <- as.data.frame(apply(randu, 2, function(x) (x-mean(x))/sd(x)))
randu_std$yz <- sqrt(35)/6*randu_std$y-randu_std$z/6

@@

Optimisation of PP is often discussed when new indexes are proposed [@posse1995; @marie-sainte2010; @grochowski2011]. @cook1995grand tied the optimisation more closely to the index, when they introduced the PP guided tour, which monitors the optimisation visually so that the user can see the projected data leading in and out of the optima. An implementation is available in the `tourr` package [@tourr] in R [@R]. @RJ-2021-105 illustrated how to diagnose optimisation processes, particularly focusing on the guided tour, and revealed a need for improved optimisation. While improving the quality of the optimisation solutions in the tour is essential, it is also important to be able to view the data projections as the optimisation progresses. Integrating the guided tour with a global optimisation algorithm that is efficient in finding the global optimal and enables viewing of the projected data during the exploration process is a goal.

Here, the potential for a Jellyfish Search Optimiser (JSO) to be
integrated with the projection pursuit guided tour is explored. JSO,
[@chou_novel_2021] - [@rajwar_exhaustive_2023], inspired by the search
behaviour of jellyfish in the ocean, is a swarm-based metaheuristic
designed to solve global optimisation problems. Compared to traditional
methods, JSO has demonstrated stronger search ability and faster
convergence, and requires fewer tuning parameters. These practical
advantages make JSO a promising candidate for enhancing PP optimisation.
<!--The study reported here explores JSO's potential in enhancing projection pursuit and examines the characteristics
and advantages of such implementation.-->

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on @barnett1981interpreting's squint
angle) and smoothness as introduced in @laa_using_2020. Here we
mathematically define metrics for squintability and smoothness, which is
new for the field. A series of simulation experiments using various
datasets and PP indexes are conducted to assess JSO's behaviour and its
sensitivity to hyper-parameter choices (number of jellyfish and maximum
number of tries). The relationship between the JSO performance,
hyper-parameter choices and properties of PP indexes (smoothness and
squintability) is analysed to provide guidance on selecting optimisers
for practitioners using projection pursuit. Additionally, it aims to
guide the design of new indexes that facilitate easy optimization for PP
researchers.

The paper is structured as follows. @sec-background introduces the
background of PP guided tour and reviews existing optimisers and index
functions in the literature. @sec-theory details JSO and introduces
metrics that measure different properties of PP indexes, smoothness and
squintability. @sec-sim-deets details two simulation experiments to
assess JSO's performance: one comparing JSO's performance improvements
relative to an existing optimiser, Creeping Random Search (CRS), and the
other studying the impact of different PP index properties on
optimisation performance. @sec-sim-res presents the results. @sec-discussion discusses implementation details and insights for practitioners. 
@sec-conclusion summarises the work and provides suggestions for future
directions.

>>>>

Projection pursuit (PP) (@kr69, @FT74, @huber85) is a dimension reduction technique aimed at identifying informative linear projections of data. This is useful for exploring high-dimensional data, and creating plots of the data that reveal the main features to use for publication. The method involves optimising an objective function known as the PP index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the criteria for what constitutes interesting or informative projections. Let $X \in \mathbb{R}^{n\times p}$ be the data matrix, $A \in\mathbb{R}^{p \times d}$ be an orthonormal matrix, where $A$ belongs to the Stiefel manifold $\mathcal{A} = V_d(\mathbb{R}^p)$. The projection $Y = XA$ is a linear transformation that maps data from a $p$-dimensional space into a $d$-dimensional space. The index function $f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}$ is a scalar function that measures an interesting aspect of the projected data, such as deviation from normality, presence of clusters, non-linear structure, or other features of interest. For a fixed sample of data, PP finds the orthonormal basis $A$ that maximises the index value of the projection, $Y = XA$:

@@

It is interesting to note that when using PP visually, one cares less about $A$ than the plane described by $A$, because the orientation in the plane is irrelevant. The space of planes belongs to a Grassmann manifold. This is usually how the projection pursuit guided tour (PPGT) [@cook1995grand] operates, when using geodesic interpolation between starting and target planes. It interpolates plane to plane, removing irrelevant within plane spin, and is agnostic to the basis ($A$) used to define the plane. Thus, indexes which are used for the PPGT should be rotationally invariant.

Index functions are quite varied in form, partially depending on the data that is being projected. @fig-example-functions shows two examples. Huber plots [@huberplot] of 2D data sets are in (a) and (c), showing the PP index values for all 1D projections of the 2D data in polar coordinates, which reveals the form of these functions. The dashed circle is a baseline set at the average value, and the straight line marks the optimal projection. Plots (b) and (d) show the respective best projections of the data as histograms. Indexes like the holes, central mass and skewness [@cook1993projection] are generally smooth for most data sets, but capture only large patterns. Many indexes noisy and non-convex, requiring an effective and efficient optimisation procedure to explore the data landscape and achieve a globally optimal viewpoint of the data. The skewness index computed for trimodal data, in (a), is smooth with a large squint angle but has three modes, and thus is not convex. The binned normality index (a simple version of a non-normality index as described in @huber85) computed on the famous RANDU data, in (c), is noisier and has a very small squint angle. The discreteness cannot be seen unless the optimiser is very close to the optimal projection.

@@

```{r}
#| label: fig-example-functions
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 80%
#| fig-cap: "Examples of PP indexes with large (top row) and small (bottom row) squint angles, shown with a Huber plot, and histogram of the projected data corresponding to the optimal projection. A Huber plot shows the PP index values for all 1D data projections in polar coordinates."
data(randu)
randu_std <- as.data.frame(apply(randu, 2, function(x) (x-mean(x))/sd(x)))
randu_std$yz <- sqrt(35)/6*randu_std$y-randu_std$z/6

@@

Optimisation of PP is often discussed when new indexes are proposed [@posse1995; @marie-sainte2010; @grochowski2011]. @cook1995grand tied the optimisation more closely to the index, when they introduced the PPGT, which monitors the optimisation visually so that the user can see the projected data leading in and out of the optima. An implementation is available in the `tourr` package [@tourr] in R [@R]. @RJ-2021-105 illustrated how to diagnose optimisation processes, particularly focusing on the guided tour, and revealed a need for improved optimisation. While improving the quality of the optimisation solutions in the tour is essential, it is also important to be able to view the data projections as the optimisation progresses. Integrating the guided tour with a global optimisation algorithm that is efficient in finding the global optimal and enables viewing of the projected data during the exploration process is a goal.

Here, the potential for a Jellyfish Search Optimiser (JSO) (see @chou_novel_2021, and @rajwar_exhaustive_2023) for the PPGT is explored. JSO, inspired by the search behaviour of jellyfish in the ocean, is a swarm-based metaheuristic designed to solve global optimisation problems. Compared to traditional methods, JSO has demonstrated stronger search ability and faster
convergence, and requires fewer tuning parameters. These practicalities
make JSO a promising candidate for enhancing PP optimisation.
<!--The study reported here explores JSO's potential in enhancing projection pursuit and examines the characteristics
and advantages of such implementation.-->

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on @barnett1981interpreting's squint
angle) and smoothness. Here, we mathematically define metrics for squintability and smoothness, which is a new contribution for PP research. A series of simulation experiments using various datasets and PP indexes are conducted to assess JSO's behaviour and its sensitivity to hyper-parameter choices (number of jellyfish and maximum number of tries). The relationship between the JSO performance, hyper-parameter choices and properties of PP indexes (smoothness and squintability) is analysed to provide guidance on selecting optimisers for practitioners using projection pursuit. Additionally, this work should guide the design of new PP indexes and facilitate better optimization for PP.

The paper is structured as follows. @sec-background introduces the
background of the PPGT, reviews existing optimisers and index
functions in the literature. @sec-theory describes JSO and introduces the 
metrics that measure different properties of PP indexes, smoothness and
squintability. @sec-sim-deets summarises the results of two simulation experiments done to assess JSO's performance: one comparing JSO's performance improvements relative to an existing optimiser, Creeping Random Search (CRS), and the other studying the impact of different PP index properties on optimisation performance. @sec-sim-res presents the results. @sec-discussion discusses implementation details and insights for practitioners. @sec-conclusion summarises the work and provides suggestions for future
directions.

### 2024-09-02 jso.tex
done revising introduction

Projection Pursuit (PP) (Kruskal (1969), Friedman and Tukey (1974),
Huber (1985)) is a dimension reduction technique aimed at identifying
informative linear projections of data. This is useful for exploring
high-dimensional data, and creating plots of the data for publication
that reveal the main features. The method involves optimising an
objective function known as the PP index (e.g. Hall (1989), Cook, Buja,
and Cabrera (1993), Lee and Cook (2010), Loperfido (2018), Loperfido
(2020)), which defines the criteria for what constitutes interesting or
informative projections. Let \(X \in \mathbb{R}^{n\times p}\) be the
data matrix, \(A \in\mathbb{R}^{p \times d}\) be an orthonormal matrix,
where \(A\) belongs to the Stiefel manifold
\(\mathcal{A} = V_d(\mathbb{R}^p)\). (It is interesting to note that
when using PP visually, one cares less about \(A\) than the plane
described by \(A\), because the orientation in the plane is irrelevant.
The the space of planes belongs to a Grassmann manifold.) The projection
\(Y = XA\) is a linear transformation that maps data from a
\(p\)-dimensional space into a \(d\)-dimensional space. The index
function \(f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}\) is a scalar
function that measures an interesting aspect of the projected data, such
as deviation from normality, presence of clusters, non-linear structure,
or other features of interest. For a fixed data, PP finds the
orthonormal basis \(A\) that maximises the index value of the
projection, \(Y = XA\):

@@

The index functions vary in structure, partly depending on the data that
is being projected. Figure~\ref{fig-example-functions} shows two
examples. Huber plots (Huber 1990) of 2D data sets are in (a) and (c).
The PP index values for all 1D projections of the 2D data are shown in
polar coordinates, revealing the nature of the function. The dashed
circle is a baseline set at the average value, and the straight line
marks the optimal projection. Plots (b) and (d) show the respective best
projections of the data as histograms. Indexes like the holes, central
mass and skewness (Cook, Buja, and Cabrera 1993) are generally smooth
for most data sets. Many indexes noisy and non-convex, requiring an
effective and efficient optimisation procedure to explore the data
landscape and achieve a globally optimal viewpoint of the data. The
skewness index computed for trimodal data, in (a), is smooth with a
large squint angle but has three modes, and thus is not convex. The
binned normality index computed on the famous RANDU data, in (c), is
noisier and has a very small squint angle. The discreteness cannot be
seen unless the optimiser is very close to the optimal projection.

@@

\caption{\label{fig-example-functions}Examples of PP indexes with large
(top row) and small (bottom row) squint angles, shown with a Huber plot,
and histogram of the projected data corresponding to the optimal
projection. Huber plot shows PP index values for all 1D projections in
polar coordinates.}

@@

Optimisation of PP is often discussed when new indexes are proposed
(Posse 1995a; Marie-Sainte, Berro, and Ruiz-Gazen 2010; Grochowski and
Duch 2011). Cook et al. (1995) tied the optimisation more closely to the
index, when they introduced the PP guided tour, which monitors the
optimisation visually so that the user can see the projected data
leading in and out of the optima. An implementation is available in the
\texttt{tourr} package (Wickham et al. 2011) in R (R Core Team 2023).
Zhang et al. (2021) illustrated how to diagnose optimisation processes,
particularly focusing on the guided tour, and revealed a need for
improved optimisation. While improving the quality of the optimisation
solutions in the tour is essential, it is also important to be able to
view the data projections as the optimisation progresses. Integrating
the guided tour with a global optimisation algorithm that is efficient
in finding the global optimal and enables viewing of the projected data
during the exploration process is a goal.

Here, the potential for a Jellyfish Search Optimiser (JSO) to be
integrated with the projection pursuit guided tour is explored. JSO,
(Chou and Truong 2021) - (Rajwar, Deep, and Das 2023), inspired by the
search behaviour of jellyfish in the ocean, is a swarm-based
metaheuristic designed to solve global optimisation problems. Compared
to traditional methods, JSO has demonstrated stronger search ability and
faster convergence, and requires fewer tuning parameters. These
practical advantages make JSO a promising candidate for enhancing PP
optimisation.

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on Tukey and Tukey (1981)'s squint angle)
and smoothness as introduced in Laa and Cook (2020). Here we
mathematically define metrics for squintability and smoothness, which is
new for the field. A series of simulation experiments using various
datasets and PP indexes are conducted to assess JSO's behaviour and its
sensitivity to hyper-parameter choices (number of jellyfish and maximum
number of tries). The relationship between the JSO performance,
hyper-parameter choices and properties of PP indexes (smoothness and
squintability) is analysed to provide guidance on selecting optimisers
for practitioners using projection pursuit. Additionally, it aims to
guide the design of new indexes that facilitate easy optimization for PP
researchers.

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes, smoothness and squintability.
Section~\ref{sec-sim-deets} details two simulation experiments to assess
JSO's performance: one comparing JSO's performance improvements relative
to an existing optimiser, Creeping Random Search (CRS), and the other
studying the impact of different PP index properties on optimisation
performance. Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-discussion} discusses implementation details and
insights for practitioners. Section~\ref{sec-conclusion} summarises the
work and provides suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, toptitle=1mm, rightrule=.15mm, colback=white, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, coltitle=black, arc=.35mm, bottomrule=.15mm, toprule=.15mm, titlerule=0mm, bottomtitle=1mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, breakable, opacitybacktitle=0.6]

>>>>

Projection pursuit (PP) (Kruskal (1969), Friedman and Tukey (1974),
Huber (1985)) is a dimension reduction technique aimed at identifying
informative linear projections of data. This is useful for exploring
high-dimensional data, and creating plots of the data that reveal the
main features to use for publication. The method involves optimising an
objective function known as the PP index (e.g. Hall (1989), Cook, Buja,
and Cabrera (1993), Lee and Cook (2010), Loperfido (2018), Loperfido
(2020)), which defines the criteria for what constitutes interesting or
informative projections. Let \(X \in \mathbb{R}^{n\times p}\) be the
data matrix, \(A \in\mathbb{R}^{p \times d}\) be an orthonormal matrix,
where \(A\) belongs to the Stiefel manifold
\(\mathcal{A} = V_d(\mathbb{R}^p)\). The projection \(Y = XA\) is a
linear transformation that maps data from a \(p\)-dimensional space into
a \(d\)-dimensional space. The index function
\(f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}\) is a scalar function
that measures an interesting aspect of the projected data, such as
deviation from normality, presence of clusters, non-linear structure, or
other features of interest. For a fixed sample of data, PP finds the
orthonormal basis \(A\) that maximises the index value of the
projection, \(Y = XA\):

@@

It is interesting to note that when using PP visually, one cares less
about \(A\) than the plane described by \(A\), because the orientation
in the plane is irrelevant. The space of planes belongs to a Grassmann
manifold. This is usually how the projection pursuit guided tour (PPGT)
(Cook et al. 1995) operates, when using geodesic interpolation between
starting and target planes. It interpolates plane to plane, removing
irrelevant within plane spin, and is agnostic to the basis (\(A\)) used
to define the plane. Thus, indexes which are used for the PPGT should be
rotationally invariant.

Index functions are quite varied in form, partially depending on the
data that is being projected. Figure~\ref{fig-example-functions} shows
two examples. Huber plots (Huber 1990) of 2D data sets are in (a) and
(c), showing the PP index values for all 1D projections of the 2D data
in polar coordinates, which reveals the form of these functions. The
dashed circle is a baseline set at the average value, and the straight
line marks the optimal projection. Plots (b) and (d) show the respective
best projections of the data as histograms. Indexes like the holes,
central mass and skewness (Cook, Buja, and Cabrera 1993) are generally
smooth for most data sets, but capture only large patterns. Many indexes
noisy and non-convex, requiring an effective and efficient optimisation
procedure to explore the data landscape and achieve a globally optimal
viewpoint of the data. The skewness index computed for trimodal data, in
(a), is smooth with a large squint angle but has three modes, and thus
is not convex. The binned normality index (a simple version of a
non-normality index as described in Huber (1985)) computed on the famous
RANDU data, in (c), is noisier and has a very small squint angle. The
discreteness cannot be seen unless the optimiser is very close to the
optimal projection.

@@

\caption{\label{fig-example-functions}Examples of PP indexes with large
(top row) and small (bottom row) squint angles, shown with a Huber plot,
and histogram of the projected data corresponding to the optimal
projection. A Huber plot shows the PP index values for all 1D data
projections in polar coordinates.}

@@

Optimisation of PP is often discussed when new indexes are proposed
(Posse 1995a; Marie-Sainte, Berro, and Ruiz-Gazen 2010; Grochowski and
Duch 2011). Cook et al. (1995) tied the optimisation more closely to the
index, when they introduced the PPGT, which monitors the optimisation
visually so that the user can see the projected data leading in and out
of the optima. An implementation is available in the \texttt{tourr}
package (Wickham et al. 2011) in R (R Core Team 2023). Zhang et al.
(2021) illustrated how to diagnose optimisation processes, particularly
focusing on the guided tour, and revealed a need for improved
optimisation. While improving the quality of the optimisation solutions
in the tour is essential, it is also important to be able to view the
data projections as the optimisation progresses. Integrating the guided
tour with a global optimisation algorithm that is efficient in finding
the global optimal and enables viewing of the projected data during the
exploration process is a goal.

Here, the potential for a Jellyfish Search Optimiser (JSO) (see Chou and
Truong (2021), and Rajwar, Deep, and Das (2023)) for the PPGT is
explored. JSO, inspired by the search behaviour of jellyfish in the
ocean, is a swarm-based metaheuristic designed to solve global
optimisation problems. Compared to traditional methods, JSO has
demonstrated stronger search ability and faster convergence, and
requires fewer tuning parameters. These practicalities make JSO a
promising candidate for enhancing PP optimisation.

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on Tukey and Tukey (1981)'s squint angle)
and smoothness. Here, we mathematically define metrics for squintability
and smoothness, which is a new contribution for PP research. A series of
simulation experiments using various datasets and PP indexes are
conducted to assess JSO's behaviour and its sensitivity to
hyper-parameter choices (number of jellyfish and maximum number of
tries). The relationship between the JSO performance, hyper-parameter
choices and properties of PP indexes (smoothness and squintability) is
analysed to provide guidance on selecting optimisers for practitioners
using projection pursuit. Additionally, this work should guide the
design of new PP indexes and facilitate better optimization for PP.

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of the PPGT, reviews existing optimisers and
index functions in the literature. Section~\ref{sec-theory} describes
JSO and introduces the metrics that measure different properties of PP
indexes, smoothness and squintability. Section~\ref{sec-sim-deets}
summarises the results of two simulation experiments done to assess
JSO's performance: one comparing JSO's performance improvements relative
to an existing optimiser, Creeping Random Search (CRS), and the other
studying the impact of different PP index properties on optimisation
performance. Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-discussion} discusses implementation details and
insights for practitioners. Section~\ref{sec-conclusion} summarises the
work and provides suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, left=2mm, rightrule=.15mm, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, breakable, opacitybacktitle=0.6, colback=white, colframe=quarto-callout-note-color-frame, toprule=.15mm, toptitle=1mm, arc=.35mm, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black]
